Sequential algorithms to split and merge terabyte-scale 3D images

### 2017-06-21 paper.tex
Updates in section II.F

% packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsfonts} % for R symbol (the set of real numbers)
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{hyperref}
\usepackage{mathtools}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

@@

As detailed in Algorithm~\ref{algo:cluster-reads}, cluster reads load
multiple blocks in memory, concatenate them in a buffer and write the
buffer sequentially in the reconstructed image. The buffer is a
slightly complex data structure, for instance an associative array or
a Python dictionary, capable of storing multiple disjoint sequences of
contiguous bytes without having to allocate memory for the bytes
between such sequences. Figure~\ref{fig:cluster-reads-buffer}
illustrates how the buffer would fill up when two blocks of one slice
and three rows per slice are loaded. Implementation details about the
buffer are given later in the document.
\begin{figure}
\centering
\def\svgwidth{\columnwidth}
\input{figures/buffer-cluster-reads.pdf_tex}
\caption{Illustration of the buffer used in cluster reads. White
  portions are not allocated. [a,b] refers to block a, row b.}
\label{fig:cluster-reads-buffer}
\end{figure}
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}
    \STATE list = sort blocks by increasing (x,y,z) values
    \STATE initialize buffer
    \STATE index = 0
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m}
        \STATE block = list[index]
        \STATE read block and insert it in buffer
        \STATE index += 1
        \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

The number of seeks performed by cluster reads depends on how blocks
loaded in memory arrange in the reconstructed image. In the best case,
complete contiguous slices of the reconstructed image can be assembled
in memory and written without seeking at all. In the worst case, some
rows in the reconstructed image are not covered by blocks loaded in
memory: $d^2$ seeks are then required, one for every partial row in
every partial slice. In the intermediary case, rows are complete but
some slices can only be partially reconstructed: $d$ seeks are then
required. See also Figure~\ref{fig:cluster-reads}. In summary:
\begin{equation}
  N_\mathrm{cluster\_reads}= n + \ceil*{\frac{bR}{m}}f(m,n,R)
\end{equation}
where:
\begin{equation}
  f(m,n,R)=
  \begin{cases}
    d^2 & $if there is an incomplete row$\\
    d   & $if there is an incomplete slice$\\
        & $but not incomplete row$\\
    1   & $otherwise$
\end{cases}
\end{equation}
\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/case1.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case2.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case3-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case3-b.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case4.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case5.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case6-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case6-b.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case7.pdf_tex}

\caption{Cluster reads configurations leading to different number of
  seeks.}
\label{fig:cluster-reads}
% See source files in
% https://drive.google.com/drive/folders/0BzT7pbvPDUdQRXppVlhfdS0ycGc?usp=sharing
\end{figure}

If $\frac{Rb}{\sqrt[3]{n}}>m\geq\frac{Rb}{\sqrt[3]{n}^2}$, i.e., a
full row of blocks fits in memory but a full slice does not, then a
variant of cluster reads consists in adjusting the memory load to the
highest value that makes sure that only full rows are written. In this
case:
\begin{equation}
  N_\mathrm{cluster\_reads\_v1}= n + \frac{bR}{m|}d
\end{equation}
where $m|=\frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}}$

And if $m\geq\frac{Rb}{\sqrt[3]{n}}$, i.e., a full slice of blocks
fits in memory, then a second variant adjusts
the memory load to the highest value that makes sure that only full
slices are written. In this case:
\begin{equation}
  N_\mathrm{cluster\_reads\_v2}= n + \frac{bR}{m||}d
\end{equation}
where $m||=\frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}$

>>>>

% packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsfonts} % for R symbol (the set of real numbers)
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{multirow}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

@@

As detailed in Algorithm~\ref{algo:cluster-reads}, cluster reads load
multiple blocks in memory, concatenate them in a buffer and write the
buffer sequentially in the reconstructed image. The buffer is a
slightly complex data structure, for instance an associative array or
a Python dictionary, capable of storing multiple disjoint sequences of
contiguous bytes without having to allocate memory for the bytes
between such sequences. Figure~\ref{fig:cluster-reads-buffer}
illustrates how the buffer would fill up when two blocks of one slice
and three rows per slice are loaded. Implementation details about the
buffer are given later.
\begin{figure}
\centering
\def\svgwidth{\columnwidth}
\input{figures/buffer-cluster-reads.pdf_tex}
\caption{Illustration of the buffer used in cluster reads. White
  portions are not allocated. [a,b] refers to block a, row b.}
\label{fig:cluster-reads-buffer}
\end{figure}
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}
    \STATE list = sort blocks by increasing (x,y,z) values
    \STATE initialize buffer
    \STATE index = 0
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m}
        \STATE block = list[index]
        \STATE read block and insert it in buffer
        \STATE index += 1
        \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

The number of seeks performed by cluster reads depends on how blocks
loaded in memory arrange in the reconstructed image. In the best case,
complete contiguous slices of the reconstructed image can be assembled
in memory and written without any seeking. In the worst case, some
rows in the reconstructed image are not covered by blocks loaded in
memory: O($d^2$) seeks are then required to write the memory load, one
for every partial row in every partial slice. In the intermediary
case, rows are complete but some slices can only be partially
reconstructed: O($d$) seeks are then
required. We have:
\begin{equation}
  N_\mathrm{cluster\_reads}= n + \sum_{i=1}^{\ceil*{\frac{bR}{m}}}f_i
\end{equation}
where $f_i$ is the number of seeks done in the $i^{th}$ memory
load. Figure~\ref{fig:cluster-reads} lists the possible block
arrangements and the corresponding values of $f_i$. \todo{Val\'erie: Check that
  ordering is consistent with Nifti's.}
%% In case (1) on
%% this Figure, the memory load completely fills rows and slices in the
%% reconstructed image. Thus it can be written with a single seek,
%% $f_i$=1. In case (2), all the rows are complete but $d$ slices are
%% incomplete; one seek is required for each slices, $f_i=d$. In case
%% (3), $d$ rows of $d$ slices are incomplete, $f_i=d^2$. In case (4),
%% $2d$ rows of $d$ slices are incomplete, $f_i=2d^2-d$. In case (5),
%% $2d$ slices are incomplete but no row is incomplete, $f_i=2d$. In case
%% (6), $d$ rows of $d$ slices are incomplete, and $d$ additional slices
%% are incomplete, $f_i$=$d^2+d-1$. And in case (7), $d$ rows of $2d$
%% slices are incomplete, $f_i$=$2d^2-1$.
In general:
\begin{equation}
  N_\mathrm{cluster\_reads}= n + \ceil*{\frac{bR}{m}}O(d^2)
\end{equation}

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/case1.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case2.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case3-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case3-b.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case4.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case5.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case6-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case6-b.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case7.pdf_tex}

\caption{In-memory block arrangements leading to different number of
  seeks in cluster reads. Blue blocks need seeking before each of
  their slices ($d$ seeks in total). Red blocks need seeking before
  each of their rows ($d^2$ seeks). Purple blocks require seeking
  before each row and each slice ($d^2$ seeks). Green blocks are
  written with no seeking.}
\label{fig:cluster-reads}
% See source files in
% https://drive.google.com/drive/folders/0BzT7pbvPDUdQRXppVlhfdS0ycGc?usp=sharing
\end{figure}

If $m\geq\frac{Rb}{\sqrt[3]{n}^2}$, i.e., a
full row of blocks fits in memory, then a
variant adjusts the memory load to the
highest value for which only full rows are written. In this
case, $f_i \in \{1,d,2d\}$ and:
\begin{equation}
  N_\mathrm{cluster\_reads\_v1}= n + \frac{bR}{m|}O(d)
\end{equation}
where
$m|=\frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}}$. It
should be noted that $m|$ is lower than $m$, therefore
$N_\mathrm{cluster\_reads\_v1}$ might be greater than
$N_\mathrm{cluster\_reads}$ for small values of $d$.

If $m\geq\frac{Rb}{\sqrt[3]{n}}$, i.e., a full slice of blocks
fits in memory, then a second variant adjusts
the memory load to the highest value for which only full
slices are written. In this case:
\begin{equation}
  N_\mathrm{cluster\_reads\_v2}= n + \frac{bR}{m||}
\end{equation}
where $m||=\frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}$.
